Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-03-25 paper/paper.tex

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is unfortunate since code
extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, most uses of extraction
(1) do not preserve the recursive structure of common implementations; and (2)
lead to unsafe casts like \ocaml{Obj.magic} in the generated code. This latter
is also problematic in that, for higher-order programs, simple interoperations
can lead to incorrect behaviour or even segfaults~\cite{forster:hal-04329663}
and, moreover, it invalidates the fast-and-loose
principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code. The
full mechanisation can be found on
Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While
programmers still need to reason about the termination of their programs, our
mechanisation will allow the use of the algebraic laws of hylomorphisms for
program reasoning, as well as extracting to idiomatic code (see
extracted \ocaml{qsort} program in Section~\ref{sec:examples}).

% As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}

>>>>

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is unfortunate since code
extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, in Abreu et al.'s approach, extraction
(1) do not preserve the recursive structure of common implementations; and (2)
lead to unsafe casts like \ocaml{Obj.magic} in the generated code. This latter
is also problematic in that, for higher-order programs, simple interoperations
can lead to incorrect behaviour or even segfaults~\cite{forster:hal-04329663}
and, moreover, it invalidates the fast-and-loose
principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation \emph{hylomorphisms} that (1) is
\emph{axiom-free} and (2) allows the extraction of idiomatic OCaml code. The
full mechanisation can be found on
Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While
programmers still need to reason about the termination of their programs, our
mechanisation will allow the use of the algebraic laws of hylomorphisms for
program reasoning, as well as extracting to idiomatic code (see
extracted \ocaml{qsort} program in Section~\ref{sec:examples}).
% As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}
